---
title: "Lab 4"
author: "Isabella Villanueva"
format: html
editor: visual
embed-sources: TRUE
---

## 1. Read in the data

*First download and then read in the met data*

```{r}
if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")
```

## 2. Prepare the data

*Remove temperatures less than -17C
Make sure there are no missing data in the key variables coded as 9999, 999, etc*
```{r}
met <- met[met$temp > -17][elev == 9999.0, elev := NA]
met
```
*Generate a date variable using the functions as.Date() (hint: You will need the following to create a date paste(year, month, day, sep = "-")).*

(need to) Using the data.table::week function, keep the observations of the first week of the month.
```{r}
met[, date := as.Date(paste(year, month, day, sep = "-"))]
met_first_week <- met[day <= 7]
```

*Compute the mean by station of the variables temp, rh, wind.sp, vis.dist, dew.point, lat, lon, and elev.*
```{r}
met_first_week <- met[,.(
  temp     = mean(temp,na.rm=TRUE),
  rh       = mean(rh,na.rm=TRUE),
  wind.sp  = mean(wind.sp,na.rm=TRUE),
  vis.dist = mean(vis.dist,na.rm=TRUE),
  lat      = mean(lat),
  lon      = mean(lon), 
  elev     = mean(elev,na.rm=TRUE)
), by=c("USAFID", "day")]
```


*Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees*
```{r}
met[, region := fifelse(lon < -98.00 & lat > 39.71, "NW",
                fifelse(lon < -98.00 & lat <= 39.71, "SW",
                fifelse(lon >= -98.00 & lat > 39.71, "NE",
                "SE")))]
```

*Create a categorical variable for elevation as in the lecture slides*
```{r}
met_avg[, elev_cat := fifelse(elev > 252, "high", "low")]
```

## 3. Use geom_violin to examine the wind speed and dew point by region
*Wind Speed*
```{r}
ggplot(met, aes(x = 1, y = wind.sp)) +
  geom_violin(fill = "pink") +
labs(title = "Wind Speed by Region",
  x = "Region",
  y = "Wind Speed") +
theme_minimal()
```

*Dew Point*
```{r}
ggplot(met, aes(x = 1, y = dew.point)) +
geom_violin(fill = "lightblue") +
labs(title = "Dew Point by Region",
x = "Region",
y = "Dew Point") +
theme_minimal()
```
